---
title: "Evaluation of performance of PTRS/OTRS when incorporating clinical risk factors"
author: "Padma Sheila Rajagopal, MD MPH"
date: "1/20/2020"

site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    df_print: paged
    toc: true
    theme: spacelab
    highlight: textmate
---
```{r DATA setup, warning=FALSE, include=FALSE, echo=FALSE}
rm(list = ls())
library(tidyverse)
library(dplyr)
library(ggplot2)
library(data.table)
library(pander)
library(ggExtra)
library(survival)
library(survminer)
library(stringr)
library(glmnet)
library(survival)
library(factoextra)
library(flashClust)
knitr::opts_chunk$set(autodep = TRUE)
options(datatable.fread.datatable = F)
options(stringsAsFactors = F)
panderOptions('table.split.table', Inf)
predixcan_cutoffs = c(5e-8, 1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1)

setwd("~/Research-Local/2020-ptrsotrs")
load("Input/PTRSOTRS.RData")


gen.score <- function(df, x, a, b) { 
  ## extract the objective column
  df1 = df[,a:b]
  col = df[, x]
  df1 * col
}

calculate_trs = function(expression_df, effect_size_df, pvalue_cutoffs = c(5e-8, 1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1)) {
  # expression_df should have rownames as gene id and colnames as individual id
  # effect_size_df should have 
  #     1. column named gene with gene id;
  #     2. column named effect_size as effect size;
  #     3. column named pvalue as pvalue
  
  # first of all, we remove genes with crazily large effect size (they will skew the PTRS distribution)
  effect_size_df_cleaned = effect_size_df %>% filter(abs(effect_size) < 5)
  
  # limit analysis to genes occcur in both expression_df and effect_size_df
  pred_expr_gene_pool = rownames(expression_df)
  predixcan_gene_pool = effect_size_df_cleaned$gene
  gene_pool_intersect = intersect(pred_expr_gene_pool, predixcan_gene_pool)
  
  # loop over p-value cutoffs and calculate PTRS at each cutoff
  out = list()
  outgenes = list()
  for(c in pvalue_cutoffs) {
    predixcan_i = effect_size_df_cleaned %>% filter(gene %in% gene_pool_intersect)
    beta_gene = predixcan_i %>% filter(pvalue < c) %>% select(gene, effect_size)
    outgenes[[length(outgenes) + 1]] = beta_gene %>% mutate(cutoff = c)
    indiv = colnames(expression_df) 
    mat = t(expression_df)
    mat = mat[, colnames(mat) %in% beta_gene$gene, drop = F]
    mat = mat[, match(beta_gene$gene, colnames(mat)), drop = F]
    tprs_i = as.matrix(mat) %*% beta_gene$effect_size
    df_tprs = data.frame(indiv = indiv, tprs = tprs_i, cutoff = c)
    rownames(df_tprs) = NULL
    out[[length(out) + 1]] = df_tprs
  }
  out = do.call(rbind, out)
  outgenes = do.call(rbind, outgenes)
  list(ptrs = out, selected_genes = outgenes)
}
```

```{r PRS score generation, cache=TRUE, warning=FALSE, echo=FALSE}
setwd("~/Research-Local/2020-ptrsotrs")
PRSChr1allpts <- read.table("Input/PRS-dosage/BRCA_chr1.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr2allpts <- read.table("Input/PRS-dosage/BRCA_chr2.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr3allpts <- read.table("Input/PRS-dosage/BRCA_chr3.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr4allpts <- read.table("Input/PRS-dosage/BRCA_chr4.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr5allpts <- read.table("Input/PRS-dosage/BRCA_chr5.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr6allpts <- read.table("Input/PRS-dosage/BRCA_chr6.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr7allpts <- read.table("Input/PRS-dosage/BRCA_chr7.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr8allpts <- read.table("Input/PRS-dosage/BRCA_chr8.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr9allpts <- read.table("Input/PRS-dosage/BRCA_chr9.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr10allpts <- read.table("Input/PRS-dosage/BRCA_chr10.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr11allpts <- read.table("Input/PRS-dosage/BRCA_chr11.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr12allpts <- read.table("Input/PRS-dosage/BRCA_chr12.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr13allpts <- read.table("Input/PRS-dosage/BRCA_chr13.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr14allpts <- read.table("Input/PRS-dosage/BRCA_chr14.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr15allpts <- read.table("Input/PRS-dosage/BRCA_chr15.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr16allpts <- read.table("Input/PRS-dosage/BRCA_chr16.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr17allpts <- read.table("Input/PRS-dosage/BRCA_chr17.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr18allpts <- read.table("Input/PRS-dosage/BRCA_chr18.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr19allpts <- read.table("Input/PRS-dosage/BRCA_chr19.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr20allpts <- read.table("Input/PRS-dosage/BRCA_chr20.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr21allpts <- read.table("Input/PRS-dosage/BRCA_chr21.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)
PRSChr22allpts <- read.table("Input/PRS-dosage/BRCA_chr22.711EUR-313PRS.dosages.tsv", sep="\t", header=TRUE)

PRSallchrallpts <- rbind(PRSChr1allpts, PRSChr2allpts, PRSChr3allpts, PRSChr4allpts, PRSChr5allpts, PRSChr6allpts, PRSChr7allpts, PRSChr8allpts, PRSChr9allpts, PRSChr10allpts, PRSChr11allpts, PRSChr12allpts, PRSChr13allpts, PRSChr14allpts, PRSChr15allpts, PRSChr16allpts, PRSChr17allpts, PRSChr18allpts, PRSChr19allpts, PRSChr20allpts, PRSChr21allpts, PRSChr22allpts) 

PRS313 <- read.csv("Input/PRS-raw/PRS-313-betas.csv", header=TRUE)
missingsnps <- as.data.frame(setdiff(PRS313$Position, PRSallchrallpts$X.2.POS))
names(missingsnps)[names(missingsnps) == "setdiff(PRS313$Position, PRSallchrallpts$X.2.POS)"] <- "missingpos"
missingsnps <- merge.data.frame(missingsnps, PRS313, by.x="missingpos", by.y="Position")

PRScolnames <- names(PRSallchrallpts)
PRScolnames <- gsub ('.*.TCGA', "TCGA", PRScolnames)
PRScolnames <- gsub(".", "-", PRScolnames, fixed = TRUE)
names(PRSallchrallpts) <- PRScolnames

PRSallchrallpts_dups <-PRSallchrallpts[duplicated(PRSallchrallpts$`X-2-POS`)|duplicated(PRSallchrallpts$`X-2-POS`, fromLast=TRUE),]
PRSallchrallpts[62,2] <- NA #Delete inappropriate duplicates
PRSallchrallpts[67,2] <- NA
PRSallchrallpts[77,2] <- NA
PRSallchrallpts[124,2] <- NA
PRSallchrallpts[174,2] <- NA

PRSallchrallpts[71,2] <- NA #Chr5, 345109
PRSallchrallpts[96,2] <- NA
PRSallchrallpts[135,2] <- NA #Chr8, 170692
PRSallchrallpts[199,2] <- NA #Chr11, 1895708
PRSallchrallpts[213,2] <- NA #Chr12, 293626
PRSallchrallpts[214,2] <- NA 
PRSallchrallpts[250,2] <- NA #Chr16, 4008542

PRSallchrallpts[92,2] <- NA #Chr5, 1279790
PRSallchrallpts[93,2] <- NA #Chr5, 1296255
PRSallchrallpts[195,2] <- NA #Chr11, 803017

PRSallchrallpts <- PRSallchrallpts %>% dplyr::filter (!is.na(PRSallchrallpts$`X-2-POS`))

PRSweights <- PRS313 %>% dplyr::select(Position, Overall.Breast.Cancer)
PRSallchrallpts <- merge(PRSallchrallpts, PRSweights, by.x="X-2-POS", by.y="Position")

PRSscore <- gen.score(PRSallchrallpts, 710, 5,709)
PRSscore[271,] <- colSums(PRSscore)
PRSscorefinal <- PRSscore[271,]
PRSscorefinal <- as.data.frame(t(PRSscorefinal))
names(PRSscorefinal)[names(PRSscorefinal) == "271"] <- "score"
PRSscorefinal$fullID <- rownames(PRSscorefinal)

PRS_sd <- sd(PRSscorefinal$score)*sqrt((length(PRSscorefinal$score)-1)/(length(PRSscorefinal$score)))
PRS_mean <- mean(PRSscorefinal$score)
PRSscorefinal$Z_score <- (PRSscorefinal$score - PRS_mean)/(PRS_sd)
PRSscorefinal$pval <- 1-pnorm(PRSscorefinal$score, PRS_mean, PRS_sd)
PRSscorefinal$cutoff <- cut(PRSscorefinal$pval, breaks=predixcan_cutoffs, include.lowest = TRUE)

PRSscorefinalOS <- merge(PRSscorefinal, famOS, by.x="fullID", by.y="submitter_id")
PRSscorefinalOS <- PRSscorefinalOS %>% dplyr::select (fullID, score, Z_score, pval, cutoff, phenotype)

PRSscorefinalPFS <- merge(PRSscorefinal, famPFS, by.x="fullID", by.y="submitter_id")
PRSscorefinalPFS <- PRSscorefinalPFS %>% dplyr::select (fullID, score, Z_score, pval, cutoff, phenotype)

survivalwhite <- merge(PRSscorefinalOS, TCGAavail)
survivalwhite$phenotype <- ifelse(survivalwhite$phenotype=="2", "case", survivalwhite$phenotype)
survivalwhite$phenotype <- ifelse(survivalwhite$phenotype=="1", 0, survivalwhite$phenotype)
survivalwhite$phenotype <- ifelse(survivalwhite$phenotype=="case", 1, survivalwhite$phenotype)
```

```{r Predicted-TPRS score generation for overall survival, cache=TRUE, warning=FALSE, echo=FALSE, results=FALSE}
setwd("~/Research-Local/2020-ptrsotrs")
ospredictedexpression <- read.table("Input/predicted_expression/BRCA/os-results_predicted_expression.txt", header=TRUE, as.is = TRUE)
ospredictedexpressionreference <- ospredictedexpression
PTRSOSgroup <- ospredictedexpressionreference$IID

PTPRSweights <- read.table("Input/risk_weights/2020_BRCA_risk_weights/BCAC_Overall_2020_spredixcan_GTExV8_mashr.csv", header=TRUE, sep=",")
PTPRSweights <- PTPRSweights %>% dplyr::select(gene, gene_name, effect_size, pvalue)

ospredictedexpression <- as.data.frame(t(ospredictedexpression))
names(ospredictedexpression) <- ospredictedexpression["IID",]

ospredictedexpression <- ospredictedexpression[-1,] #Removing the FID row
ospredictedexpression <- ospredictedexpression[-1,] #Removing the IID row 
ospredictedexpressiongene <- rownames(ospredictedexpression)
ospredictedexpression <- as.data.frame(lapply(ospredictedexpression, as.numeric))
rownames(ospredictedexpression) <- ospredictedexpressiongene
ospredictedexpression$gene <- row.names(ospredictedexpression)

ospredictedexpression <- merge(ospredictedexpression, PTPRSweights, by="gene", all.x=TRUE)

PTRSscoreraw <- gen.score(ospredictedexpression, 697, 2, 695)
PTRSscoredev <- PTRSscoreraw
rownames(PTRSscoredev) <- ospredictedexpressiongene

PTRSoptimization <- calculate_trs(PTRSscoredev, PTPRSweights)
pheno <- famOS %>% select(fullID,phenotype)
names(pheno) = c("SUBJID", "phenotype") 

perf_t = list()
ptrs <- PTRSoptimization$ptrs
ptrs$indiv = stringr::str_replace_all(ptrs$indiv, '\\.', '-')
for(c in predixcan_cutoffs) {
  perf_t[[as.character(c)]] = inner_join(pheno, ptrs %>% dplyr::filter(cutoff == c), by = c('SUBJID' = 'indiv'))
}

for(i in names(perf_t)) {
    p = perf_t[[i]] %>% ggplot() + geom_density(aes(x = tprs, fill = factor(phenotype)), alpha = .5) + theme(legend.position = 'bottom')
    p = p + ggtitle(paste0('p-value cutoff: ', i))
    ggsave(paste0('~/Desktop/', 'ptrs_cutoff_', i, '.pdf'), p, height = 10, width = 6)
  }

PTRSscorefinal <- PTRSoptimization$ptrs %>% dplyr::filter(cutoff==1e-5)
PTRSscorefinal$indiv = stringr::str_replace_all(PTRSscorefinal$indiv, '\\.', '-')
PTRSscorefinal <- merge(PTRSscorefinal, famOS, by.x="indiv", by.y="fullID")
PTRSscorefinal <- PTRSscorefinal %>% select (indiv, tprs, cutoff, phenotype)

survivalwhitePTRS <- merge(PTRSscorefinal, TCGAavail, by.x="indiv", by.y="submitter_id")
survivalwhitePTRS$phenotype <- ifelse(survivalwhitePTRS$phenotype=="2", "case", survivalwhitePTRS$phenotype)
survivalwhitePTRS$phenotype <- ifelse(survivalwhitePTRS$phenotype=="1", 0, survivalwhitePTRS$phenotype)
survivalwhitePTRS$phenotype <- ifelse(survivalwhitePTRS$phenotype=="case", 1, survivalwhitePTRS$phenotype)
```

# OTRS-Overall survival in breast cancer 
```{r Observed-TPRS score generation, cache=TRUE, warning=FALSE, echo=FALSE, results=FALSE}
setwd("~/Research-Local/2020-ptrsotrs")
osobservedexpression <- read.table("Input/observed_expression/BRCA-obexp/TCGA-BRCA_mRNA.csv", header=TRUE, sep=",")

samplesEURmatch <- famOS$fullID

osobservedexpression$submitter_id_long <- substr(osobservedexpression$aliquot_barcode, 1, 16)
osobservedexpression$submitter_id <- substr(osobservedexpression$aliquot_barcode, 1, 12)
osobservedexpression_cleaned = osobservedexpression %>% dplyr::filter(osobservedexpression$submitter_id %in% samplesEURmatch)
osobservedexpression_cleaned = osobservedexpression_cleaned %>% dplyr::filter(substr(osobservedexpression_cleaned$submitter_id_long, 14, 16)=="01A") #Isolates to primary tumor only, 682 samples remaining

osobservedexpression_cleaned$aliquot_barcode <- NULL

osobservedexpressiongenes <- colnames(osobservedexpression_cleaned)
rownames(osobservedexpression_cleaned) <- osobservedexpression_cleaned$submitter_id
osobservedexpressionnames <- rownames(osobservedexpression_cleaned)
osobservedexpressiongenes <- osobservedexpressiongenes[1:17321]

osobservedexpression_cleaned <- as.data.frame(t(osobservedexpression_cleaned))
osobservedexpression_cleaned <- osobservedexpression_cleaned[1:17321,] #removes text rows and only expression data observed
osobservedexpression_cleaned <- as.data.frame(lapply(osobservedexpression_cleaned, as.numeric))
osobservedexpressiongenes <- osobservedexpressiongenes[1:17321]
rownames(osobservedexpression_cleaned) <- osobservedexpressiongenes

OTPRSweights <- read.table("Input/risk_weights/2020_BRCA_risk_weights/BCAC_Overall_2020_spredixcan_GTExV8_mashr.csv", header=TRUE, sep=",")
OTPRSweights <- OTPRSweights %>% select(gene, gene_name, effect_size, pvalue)

osobservedexpression_cleaned$gene_name <- row.names(osobservedexpression_cleaned)
osobservedexpression_reference <- osobservedexpression_cleaned 

osobservedexpression_cleaned <- merge(osobservedexpression_cleaned, OTPRSweights, by="gene_name", all.x=TRUE)
osobservedexpression_cleaned <- osobservedexpression_cleaned %>% dplyr::filter(!is.na(effect_size))
osobservedexpression_cleaned <- osobservedexpression_cleaned %>% dplyr::filter(duplicated(gene_name)==FALSE)

osobservedexpressiongenes <- osobservedexpression_cleaned$gene
osobservedexpressionrows <- osobservedexpression_cleaned$gene_name

osobservedexpression_cleaned_t <- osobservedexpression_cleaned %>%
  mutate_at(vars(-gene_name, -gene, -effect_size, -pvalue), function(x) {log(x+1)})

#OTRSscore <- gen.score(observedexpression_cleaned, 647, 2, 645)
OTRSscore <- gen.score(osobservedexpression_cleaned_t, 685, 2, 683)

OTRSscoredev <- OTRSscore
rownames(OTRSscoredev) <- osobservedexpressiongenes

OTRSoptimization <- calculate_trs(OTRSscoredev, OTPRSweights)

pheno <- famOS %>% select(fullID,phenotype)
names(pheno) = c("SUBJID", "phenotype") 

perf_o = list()
otrs <- OTRSoptimization$ptrs
otrs$indiv = stringr::str_replace_all(otrs$indiv, '\\.', '-')
for(c in predixcan_cutoffs) {
  perf_o[[as.character(c)]] = inner_join(pheno, otrs %>% dplyr::filter(cutoff == c), by = c('SUBJID' = 'indiv'))
}

for(i in names(perf_o)) {
    p = perf_o[[i]] %>% ggplot() + geom_density(aes(x = tprs, fill = factor(phenotype)), alpha = .5) + theme(legend.position = 'bottom')
    p = p + ggtitle(paste0('p-value cutoff: ', i))
    ggsave(paste0('~/Desktop/', 'otrs_cutoff_', i, '.pdf'), p, height = 10, width = 6)
  }

OTRSscorefinal <- OTRSoptimization$ptrs %>% filter(cutoff==1e-5)
OTRSscorefinal$indiv = stringr::str_replace_all(OTRSscorefinal$indiv, '\\.', '-')
OTRSscorefinal <- merge(OTRSscorefinal, famOS, by.x="indiv", by.y="fullID")
OTRSscorefinal <- OTRSscorefinal %>% select (indiv, tprs, cutoff, phenotype)

survivalwhiteOTRS <- merge(OTRSscorefinal, TCGAavail, by.x="indiv", by.y="submitter_id")
```

# Clinical: Overall survival
```{r Cox regression of combined methods and overall survival, cache=TRUE, echo=FALSE, warning=FALSE}
setwd("~/Research-Local/2020-ptrsotrs")
mergePTRS <- survivalwhitePTRS %>% dplyr::select (indiv, tprs)
colnames(mergePTRS) <- c("indiv", "PTRS")

mergeOTRS <- survivalwhiteOTRS %>% dplyr::select (indiv, tprs)
colnames(mergeOTRS) <- c("indiv", "OTRS")
                                                    
survivalwhiteall <- merge(survivalwhite, mergePTRS, by.x="submitter_id", by.y="indiv", all.y=TRUE)
survivalwhiteall <- merge(survivalwhiteall, mergeOTRS, by.x="submitter_id", by.y="indiv", all.y=TRUE)

survivalwhiteall$AJCCN <- ifelse (survivalwhiteall$AJCCN=="N0" | survivalwhiteall$AJCCN=="N0 (I-)" | survivalwhiteall$AJCCN=="N0 (I+)", 0, survivalwhiteall$AJCCN)
survivalwhiteall$AJCCN <- ifelse (survivalwhiteall$AJCCN=="N1" | survivalwhiteall$AJCCN=="N1A" | survivalwhiteall$AJCCN=="N1B" | survivalwhiteall$AJCCN=="N1MI" , 1, survivalwhiteall$AJCCN)
survivalwhiteall$AJCCN <- ifelse (survivalwhiteall$AJCCN=="N2" | survivalwhiteall$AJCCN=="N2A", 2, survivalwhiteall$AJCCN)
survivalwhiteall$AJCCN <- ifelse (survivalwhiteall$AJCCN=="N3" | survivalwhiteall$AJCCN=="N3A" | survivalwhiteall$AJCCN=="N3B", 3, survivalwhiteall$AJCCN)
survivalwhiteall$AJCCN <- ifelse (survivalwhiteall$AJCCN=="NX", NA, survivalwhiteall$AJCCN)

survivalwhiteall$AJCCT <- ifelse (survivalwhiteall$AJCCT=="T1" | survivalwhiteall$AJCCT=="T1B" | survivalwhiteall$AJCCT=="T1C", 1, survivalwhiteall$AJCCT)
survivalwhiteall$AJCCT <- ifelse (survivalwhiteall$AJCCT=="T2" | survivalwhiteall$AJCCT=="T2A", 2, survivalwhiteall$AJCCT)
survivalwhiteall$AJCCT <- ifelse (survivalwhiteall$AJCCT=="T3" | survivalwhiteall$AJCCT=="T3A", 3, survivalwhiteall$AJCCT)
survivalwhiteall$AJCCT <- ifelse (survivalwhiteall$AJCCT=="T4" | survivalwhiteall$AJCCT=="T4B" | survivalwhiteall$AJCCT=="T4D", 4, survivalwhiteall$AJCCT)
survivalwhiteall$AJCCT <- ifelse (survivalwhiteall$AJCCT=="TX", NA, survivalwhiteall$AJCCT)

survivalwhiteall$AJCCM <- ifelse (survivalwhiteall$AJCCM=="CM0 (I+)" | survivalwhiteall$AJCCM=="M0", 0, survivalwhiteall$AJCCM)
survivalwhiteall$AJCCM <- ifelse (survivalwhiteall$AJCCM=="M1", 1, survivalwhiteall$AJCCM)
survivalwhiteall$AJCCM <- ifelse (survivalwhiteall$AJCCM=="MX", NA, survivalwhiteall$AJCCM)

survivalwhiteall$er <- ifelse (survivalwhiteall$er=="Negative", 0, survivalwhiteall$er)
survivalwhiteall$er <- ifelse (survivalwhiteall$er=="Positive", 1, survivalwhiteall$er)

survivalwhiteall$pr <- ifelse (survivalwhiteall$pr=="Negative", 0, survivalwhiteall$pr)
survivalwhiteall$pr <- ifelse (survivalwhiteall$pr=="Positive", 1, survivalwhiteall$pr)

survivalwhiteall$hr <- NA
survivalwhiteall$hr <- ifelse (survivalwhiteall$er=="0" & survivalwhiteall$pr=="0", 0, survivalwhiteall$hr)
survivalwhiteall$hr <- ifelse (survivalwhiteall$er=="1" | survivalwhiteall$pr=="1", 1, survivalwhiteall$hr)

survivalwhiteall$her2 <- ifelse (survivalwhiteall$her2=="Negative", 0, survivalwhiteall$her2)
survivalwhiteall$her2 <- ifelse (survivalwhiteall$her2=="Positive", 1, survivalwhiteall$her2)

#Cox regression of survival
covariates <- c("age", "AJCCN",  "AJCCT", "AJCCM", "hr", "her2", "score", "PTRS", "OTRS")

class(survivalwhiteall$oss.months) = 'numeric'
class(survivalwhiteall$phenotype) = 'numeric'
class(survivalwhiteall$age) = 'numeric'
class(survivalwhiteall$AJCCN) = 'numeric'
class(survivalwhiteall$AJCCT) = 'numeric'
class(survivalwhiteall$AJCCM) = 'numeric'
class(survivalwhiteall$hr) <- 'numeric'
class(survivalwhiteall$her2) <- 'numeric'
class(survivalwhiteall$score) = 'numeric'
class(survivalwhiteall$PTRS) = 'numeric'
class(survivalwhiteall$OTRS) = 'numeric'

univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('Surv(survivalwhiteall$oss.months, survivalwhiteall$phenotype)~', x)))

univ_models <- lapply( univ_formulas, function(x){coxph(x, data = survivalwhiteall)})

# Extract data 
univ_results <- lapply(univ_models,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          wald.test<-signif(x$wald["test"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coeficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          HR <- paste0(HR, " (", 
                                       HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, wald.test, p.value)
                          names(res)<-c("beta", "HR (95% CI for HR)", "wald.test", 
                                        "p.value")
                          return(res)
                          #return(exp(cbind(coef(x),confint(x))))
                         })
res <- t(as.data.frame(univ_results, check.names = FALSE))
as.data.frame(res)

OS_European <- survivalwhiteall %>% select(oss.months, phenotype, age, AJCCN, AJCCT, AJCCM, hr, her2, PTRS)
colnames(OS_European) = c("oss.months", "phenotype", "Age", "AJCCN.Nodal_Status", "AJCC.Tumor_Size", "AJCC.Metastasis", "Hormone.Receptor.Pos", "HER2.Receptor.Pos", "Predicted_Transcriptomic_Score")

surv_objectall <- Surv(time=OS_European$oss.months, event=as.numeric(OS_European$phenotype))

fit.coxphall <- coxph(surv_objectall ~ OS_European$Age+OS_European$AJCCN.Nodal_Status+OS_European$AJCC.Tumor_Size+OS_European$AJCC.Metastasis+OS_European$Hormone.Receptor.Pos+OS_European$HER2.Receptor.Pos+OS_European$Predicted_Transcriptomic_Score, data=OS_European) 
ggforest(fit.coxphall, data = OS_European, main="Hazard ratios of all data in TCGA patients\n overall survival, EUR ancestry")
```

# Clinical: Progression-free survival 
```{r Cox regression of combined methods and progression-free survival, echo=FALSE, cache=TRUE, warning=FALSE, execute=FALSE}
setwd("~/Research-Local/2020-ptrsotrs")
recurrencewhite <- merge(PRSscorefinalPFS, TCGAavail)
recurrencewhite$phenotype <- ifelse(recurrencewhite$phenotype=="2", "case", recurrencewhite$phenotype)
recurrencewhite$phenotype <- ifelse(recurrencewhite$phenotype=="1", 0, recurrencewhite$phenotype)
recurrencewhite$phenotype <- ifelse(recurrencewhite$phenotype=="case", 1, recurrencewhite$phenotype)

recurrencewhitePTRS <- merge(PTRSscorefinal, TCGAavail, by.x="indiv", by.y="submitter_id")
recurrencewhitePTRS$phenotype <- ifelse(recurrencewhitePTRS$phenotype=="2", "case", recurrencewhitePTRS$phenotype)
recurrencewhitePTRS$phenotype <- ifelse(recurrencewhitePTRS$phenotype=="1", 0, recurrencewhitePTRS$phenotype)
recurrencewhitePTRS$phenotype <- ifelse(recurrencewhitePTRS$phenotype=="case", 1, recurrencewhitePTRS$phenotype)

recurrencewhiteOTRS <- merge(OTRSscorefinal, TCGAavail, by.x="indiv", by.y="submitter_id")

mergePTRS <- recurrencewhitePTRS %>% dplyr::select (indiv, tprs)
colnames(mergePTRS) <- c("indiv", "PTRS")

mergeOTRS <- recurrencewhiteOTRS %>% dplyr::select (indiv, tprs)
colnames(mergeOTRS) <- c("indiv", "OTRS")
                                                    
recurrencewhiteall <- merge(recurrencewhite, mergePTRS, by.x="submitter_id", by.y="indiv", all.y=TRUE)
recurrencewhiteall <- merge(recurrencewhiteall, mergeOTRS, by.x="submitter_id", by.y="indiv", all.y=TRUE)

recurrencewhiteall$AJCCN <- ifelse (recurrencewhiteall$AJCCN=="N0" | recurrencewhiteall$AJCCN=="N0 (I-)" | recurrencewhiteall$AJCCN=="N0 (I+)", 0, recurrencewhiteall$AJCCN)
recurrencewhiteall$AJCCN <- ifelse (recurrencewhiteall$AJCCN=="N1" | recurrencewhiteall$AJCCN=="N1A" | recurrencewhiteall$AJCCN=="N1B" | recurrencewhiteall$AJCCN=="N1MI" , 1, recurrencewhiteall$AJCCN)
recurrencewhiteall$AJCCN <- ifelse (recurrencewhiteall$AJCCN=="N2" | recurrencewhiteall$AJCCN=="N2A", 2, recurrencewhiteall$AJCCN)
recurrencewhiteall$AJCCN <- ifelse (recurrencewhiteall$AJCCN=="N3" | recurrencewhiteall$AJCCN=="N3A" | recurrencewhiteall$AJCCN=="N3B", 3, recurrencewhiteall$AJCCN)
recurrencewhiteall$AJCCN <- ifelse (recurrencewhiteall$AJCCN=="NX", NA, recurrencewhiteall$AJCCN)

recurrencewhiteall$AJCCT <- ifelse (recurrencewhiteall$AJCCT=="T1" | recurrencewhiteall$AJCCT=="T1B" | recurrencewhiteall$AJCCT=="T1C", 1, recurrencewhiteall$AJCCT)
recurrencewhiteall$AJCCT <- ifelse (recurrencewhiteall$AJCCT=="T2" | recurrencewhiteall$AJCCT=="T2A", 2, recurrencewhiteall$AJCCT)
recurrencewhiteall$AJCCT <- ifelse (recurrencewhiteall$AJCCT=="T3" | recurrencewhiteall$AJCCT=="T3A", 3, recurrencewhiteall$AJCCT)
recurrencewhiteall$AJCCT <- ifelse (recurrencewhiteall$AJCCT=="T4" | recurrencewhiteall$AJCCT=="T4B" | recurrencewhiteall$AJCCT=="T4D", 4, recurrencewhiteall$AJCCT)
recurrencewhiteall$AJCCT <- ifelse (recurrencewhiteall$AJCCT=="TX", NA, recurrencewhiteall$AJCCT)

recurrencewhiteall$AJCCM <- ifelse (recurrencewhiteall$AJCCM=="CM0 (I+)" | recurrencewhiteall$AJCCM=="M0", 0, recurrencewhiteall$AJCCM)
recurrencewhiteall$AJCCM <- ifelse (recurrencewhiteall$AJCCM=="M1", 1, recurrencewhiteall$AJCCM)
recurrencewhiteall$AJCCM <- ifelse (recurrencewhiteall$AJCCM=="MX", NA, recurrencewhiteall$AJCCM)

recurrencewhiteall$er <- ifelse (recurrencewhiteall$er=="Negative", 0, recurrencewhiteall$er)
recurrencewhiteall$er <- ifelse (recurrencewhiteall$er=="Positive", 1, recurrencewhiteall$er)

recurrencewhiteall$pr <- ifelse (recurrencewhiteall$pr=="Negative", 0, recurrencewhiteall$pr)
recurrencewhiteall$pr <- ifelse (recurrencewhiteall$pr=="Positive", 1, recurrencewhiteall$pr)

recurrencewhiteall$hr <- NA
recurrencewhiteall$hr <- ifelse (recurrencewhiteall$er=="0" & recurrencewhiteall$pr=="0", 0, recurrencewhiteall$hr)
recurrencewhiteall$hr <- ifelse (recurrencewhiteall$er=="1" | recurrencewhiteall$pr=="1", 1, recurrencewhiteall$hr)

recurrencewhiteall$her2 <- ifelse (recurrencewhiteall$her2=="Negative", 0, recurrencewhiteall$her2)
recurrencewhiteall$her2 <- ifelse (recurrencewhiteall$her2=="Positive", 1, recurrencewhiteall$her2)

recurrencewhiteall$pam50n <- recurrencewhiteall$pam50
recurrencewhiteall$pam50n <- ifelse (recurrencewhiteall$pam50n=="Normal-like", 1, recurrencewhiteall$pam50n)
recurrencewhiteall$pam50n <- ifelse (recurrencewhiteall$pam50n=="Luminal A", 0, recurrencewhiteall$pam50n)
recurrencewhiteall$pam50n <- ifelse (recurrencewhiteall$pam50n=="Luminal B", 0, recurrencewhiteall$pam50n)
recurrencewhiteall$pam50n <- ifelse (recurrencewhiteall$pam50n=="HER2-enriched", 0, recurrencewhiteall$pam50n)
recurrencewhiteall$pam50n <- ifelse (recurrencewhiteall$pam50n=="Basal-like", 0, recurrencewhiteall$pam50n)

recurrencewhiteall$pam50luma <- recurrencewhiteall$pam50
recurrencewhiteall$pam50luma <- ifelse (recurrencewhiteall$pam50luma=="Normal-like", 0, recurrencewhiteall$pam50luma)
recurrencewhiteall$pam50luma <- ifelse (recurrencewhiteall$pam50luma=="Luminal A", 1, recurrencewhiteall$pam50luma)
recurrencewhiteall$pam50luma <- ifelse (recurrencewhiteall$pam50luma=="Luminal B", 0, recurrencewhiteall$pam50luma)
recurrencewhiteall$pam50luma <- ifelse (recurrencewhiteall$pam50luma=="HER2-enriched", 0, recurrencewhiteall$pam50luma)
recurrencewhiteall$pam50luma <- ifelse (recurrencewhiteall$pam50luma=="Basal-like", 0, recurrencewhiteall$pam50luma)

recurrencewhiteall$pam50lumb <- recurrencewhiteall$pam50
recurrencewhiteall$pam50lumb <- ifelse (recurrencewhiteall$pam50lumb=="Normal-like", 0, recurrencewhiteall$pam50lumb)
recurrencewhiteall$pam50lumb <- ifelse (recurrencewhiteall$pam50lumb=="Luminal A", 0, recurrencewhiteall$pam50lumb)
recurrencewhiteall$pam50lumb <- ifelse (recurrencewhiteall$pam50lumb=="Luminal B", 1, recurrencewhiteall$pam50lumb)
recurrencewhiteall$pam50lumb <- ifelse (recurrencewhiteall$pam50lumb=="HER2-enriched", 0, recurrencewhiteall$pam50lumb)
recurrencewhiteall$pam50lumb <- ifelse (recurrencewhiteall$pam50lumb=="Basal-like", 0, recurrencewhiteall$pam50lumb)

recurrencewhiteall$pam50her2 <- recurrencewhiteall$pam50
recurrencewhiteall$pam50her2 <- ifelse (recurrencewhiteall$pam50her2=="Normal-like", 0, recurrencewhiteall$pam50her2)
recurrencewhiteall$pam50her2 <- ifelse (recurrencewhiteall$pam50her2=="Luminal A", 0, recurrencewhiteall$pam50her2)
recurrencewhiteall$pam50her2 <- ifelse (recurrencewhiteall$pam50her2=="Luminal B", 0, recurrencewhiteall$pam50her2)
recurrencewhiteall$pam50her2 <- ifelse (recurrencewhiteall$pam50her2=="HER2-enriched", 1, recurrencewhiteall$pam50her2)
recurrencewhiteall$pam50her2 <- ifelse (recurrencewhiteall$pam50her2=="Basal-like", 0, recurrencewhiteall$pam50her2)

recurrencewhiteall$pam50basal <- recurrencewhiteall$pam50
recurrencewhiteall$pam50basal <- ifelse (recurrencewhiteall$pam50basal=="Normal-like", 0, recurrencewhiteall$pam50basal)
recurrencewhiteall$pam50basal <- ifelse (recurrencewhiteall$pam50basal=="Luminal A", 0, recurrencewhiteall$pam50basal)
recurrencewhiteall$pam50basal <- ifelse (recurrencewhiteall$pam50basal=="Luminal B", 0, recurrencewhiteall$pam50basal)
recurrencewhiteall$pam50basal <- ifelse (recurrencewhiteall$pam50basal=="HER2-enriched", 0, recurrencewhiteall$pam50basal)
recurrencewhiteall$pam50basal <- ifelse (recurrencewhiteall$pam50basal=="Basal-like", 1, recurrencewhiteall$pam50basal)

#Cox regression of survival
covariates <- c("age", "AJCCN",  "AJCCT", "AJCCM", "hr", "her2", "pam50luma", "pam50lumb", "pam50her2", "pam50basal", "score", "PTRS", "OTRS")

class(recurrencewhiteall$oss.months) = 'numeric'
class(recurrencewhiteall$phenotype) = 'numeric'
class(recurrencewhiteall$age) = 'numeric'
class(recurrencewhiteall$AJCCN) = 'numeric'
class(recurrencewhiteall$AJCCT) = 'numeric'
class(recurrencewhiteall$AJCCM) = 'numeric'
class(recurrencewhiteall$hr) <- 'numeric'
class(recurrencewhiteall$her2) <- 'numeric'
class(recurrencewhiteall$pam50luma) <- 'numeric'
class(recurrencewhiteall$pam50lumb) <- 'numeric'
class(recurrencewhiteall$pam50her2) <- 'numeric'
class(recurrencewhiteall$pam50basal) <- 'numeric'
class(recurrencewhiteall$score) = 'numeric'
class(recurrencewhiteall$PTRS) = 'numeric'
class(recurrencewhiteall$OTRS) = 'numeric'

univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('Surv(recurrencewhiteall$oss.months, recurrencewhiteall$phenotype)~', x)))

univ_models <- lapply( univ_formulas, function(x){coxph(x, data = recurrencewhiteall)})

# Extract data 
univ_results <- lapply(univ_models,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          wald.test<-signif(x$wald["test"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coeficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          HR <- paste0(HR, " (", 
                                       HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, wald.test, p.value)
                          names(res)<-c("beta", "HR (95% CI for HR)", "wald.test", 
                                        "p.value")
                          return(res)
                          #return(exp(cbind(coef(x),confint(x))))
                         })
res <- t(as.data.frame(univ_results, check.names = FALSE))
as.data.frame(res)

PFS_European <- recurrencewhiteall %>% select(oss.months, phenotype, age, AJCCN, AJCCT, AJCCM, pam50luma, pam50her2, pam50basal, score, PTRS, OTRS)
colnames(PFS_European) = c("oss.months", "phenotype", "Age", "AJCCN.Nodal_Status", "AJCC.Tumor_Size", "AJCC.Metastasis", "LumA.PAM50","HER2.PAM50", "Basal.PAM50","Polygenic_Score", "Predicted_Transcriptomic_Score", "Observed_Transcriptomic_Score")
surv_objectall <- Surv(time=PFS_European$oss.months, event=as.numeric(PFS_European$phenotype))

fit.coxphall <- coxph(surv_objectall ~ PFS_European$Age+PFS_European$AJCCN.Nodal_Status+PFS_European$AJCC.Tumor_Size+PFS_European$LumA.PAM50+PFS_European$Basal.PAM50+PFS_European$HER2.PAM50+PFS_European$Polygenic_Score+PFS_European$Predicted_Transcriptomic_Score+PFS_European$Observed_Transcriptomic_Score, data=PFS_European) 

ggforest(fit.coxphall, data = PFS_European, main="Hazard ratios of all data in TCGA patients\n progression-free survival, EUR ancestry")

#Use this code to produce the graphs of hormone receptor status without scores included
PFS_European <- recurrencewhiteall %>% select(oss.months, phenotype, age, AJCCN, AJCCT, AJCCM, pam50luma, pam50lumb, pam50her2, pam50basal)
colnames(PFS_European) = c("oss.months", "phenotype", "Age", "AJCCN.Nodal_Status", "AJCC.Tumor_Size", "AJCC.Metastasis", "LumA.PAM50", "LumB.PAM50","HER2.PAM50", "Basal.PAM50")
surv_objectall <- Surv(time=PFS_European$oss.months, event=as.numeric(PFS_European$phenotype))

fit.coxphall <- coxph(surv_objectall ~ PFS_European$Age+PFS_European$AJCCN.Nodal_Status+PFS_European$AJCC.Tumor_Size+PFS_European$LumB.PAM50+PFS_European$HER2.PAM50+PFS_European$Basal.PAM50, data=PFS_European) 

ggforest(fit.coxphall, data = PFS_European, main="Hazard ratios of all data in TCGA patients\n progression-free survival, EUR ancestry")
```

When combined with clinical data, the score performance of PTRS in overall survival remained persistent, offering an interesting signal to study further. The performance of OTRS in PFS was actually improved with clinical data (more consistent with real-world data), but this was no longer significant with incorporation of PAM50 data, suggesting that the latter encompasses the same information that the former does and the germline does not add specific information in that context.
