---
title: "Generation of observed transcriptomic risk score in patients with breast cancer"
author: "Padma Sheila Rajagopal, MD MPH"
date: "1/13/2020"

site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    df_print: paged
    toc: true
    theme: spacelab
    highlight: textmate
    
---
```{r DATA setup, warning=FALSE, include=FALSE, echo=FALSE}
rm(list = ls())
library(tidyverse)
library(dplyr)
library(ggplot2)
library(data.table)
library(pander)
library(ggExtra)
library(survival)
library(survminer)
library(stringr)
library(glmnet)
library(survival)
library(factoextra)
library(flashClust)
knitr::opts_chunk$set(autodep = TRUE)
options(datatable.fread.datatable = F)
options(stringsAsFactors = F)
panderOptions('table.split.table', Inf)
predixcan_cutoffs = c(5e-8, 1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1)

setwd("~/Research-Local/2020-ptrsotrs")
load("Input/PTRSOTRS.RData")

gen.score <- function(df, x, a, b) { 
  ## extract the objective column
  df1 = df[,a:b]
  col = df[, x]
  df1 * col
}

calculate_trs = function(expression_df, effect_size_df, pvalue_cutoffs = c(5e-8, 1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1)) {
  # expression_df should have rownames as gene id and colnames as individual id
  # effect_size_df should have 
  #     1. column named gene with gene id;
  #     2. column named effect_size as effect size;
  #     3. column named pvalue as pvalue
  
  # first of all, we remove genes with crazily large effect size (they will skew the PTRS distribution)
  effect_size_df_cleaned = effect_size_df %>% filter(abs(effect_size) < 5)
  
  # limit analysis to genes occcur in both expression_df and effect_size_df
  pred_expr_gene_pool = rownames(expression_df)
  predixcan_gene_pool = effect_size_df_cleaned$gene
  gene_pool_intersect = intersect(pred_expr_gene_pool, predixcan_gene_pool)
  
  # loop over p-value cutoffs and calculate PTRS at each cutoff
  out = list()
  outgenes = list()
  for(c in pvalue_cutoffs) {
    predixcan_i = effect_size_df_cleaned %>% filter(gene %in% gene_pool_intersect)
    beta_gene = predixcan_i %>% filter(pvalue < c) %>% select(gene, effect_size)
    outgenes[[length(outgenes) + 1]] = beta_gene %>% mutate(cutoff = c)
    indiv = colnames(expression_df) 
    mat = t(expression_df)
    mat = mat[, colnames(mat) %in% beta_gene$gene, drop = F]
    mat = mat[, match(beta_gene$gene, colnames(mat)), drop = F]
    tprs_i = as.matrix(mat) %*% beta_gene$effect_size
    df_tprs = data.frame(indiv = indiv, tprs = tprs_i, cutoff = c)
    rownames(df_tprs) = NULL
    out[[length(out) + 1]] = df_tprs
  }
  out = do.call(rbind, out)
  outgenes = do.call(rbind, outgenes)
  list(ptrs = out, selected_genes = outgenes)
}
```

# Germline*Somatic: Observed transcriptomic score (OTRS)

This is the code to generate the observed transcriptomic risk score and assess its ability to prognosticate overall survival and progression free survival among breast cancer patients. 

The observed transcriptomic score is derived from using the weight of a gene's association to germline risk of breast cancer * observed expression of that gene in the tumor, summed across different thresholds of significance. 

# OTRS-Overall survival in breast cancer 
```{r Observed-TPRS score generation, cache=TRUE, warning=FALSE, echo=FALSE}
setwd("~/Research-Local/2020-ptrsotrs")
osobservedexpression <- read.table("Input/observed_expression/BRCA-obexp/TCGA-BRCA_mRNA.csv", header=TRUE, sep=",")

samplesEURmatch <- famOS$fullID

osobservedexpression$submitter_id_long <- substr(osobservedexpression$aliquot_barcode, 1, 16)
osobservedexpression$submitter_id <- substr(osobservedexpression$aliquot_barcode, 1, 12)
osobservedexpression_cleaned = osobservedexpression %>% dplyr::filter(osobservedexpression$submitter_id %in% samplesEURmatch)
osobservedexpression_cleaned = osobservedexpression_cleaned %>% dplyr::filter(substr(osobservedexpression_cleaned$submitter_id_long, 14, 16)=="01A") #Isolates to primary tumor only, 682 samples remaining

osobservedexpression_cleaned$aliquot_barcode <- NULL

osobservedexpressiongenes <- colnames(osobservedexpression_cleaned)
rownames(osobservedexpression_cleaned) <- osobservedexpression_cleaned$submitter_id
osobservedexpressionnames <- rownames(osobservedexpression_cleaned)
osobservedexpressiongenes <- osobservedexpressiongenes[1:17321]

osobservedexpression_cleaned <- as.data.frame(t(osobservedexpression_cleaned))
osobservedexpression_cleaned <- osobservedexpression_cleaned[1:17321,] #removes text rows and only expression data observed
osobservedexpression_cleaned <- as.data.frame(lapply(osobservedexpression_cleaned, as.numeric))
osobservedexpressiongenes <- osobservedexpressiongenes[1:17321]
rownames(osobservedexpression_cleaned) <- osobservedexpressiongenes

OTPRSweights <- read.table("Input/risk_weights/2020_BRCA_risk_weights/BCAC_Overall_2020_spredixcan_GTExV8_mashr.csv", header=TRUE, sep=",")
OTPRSweights <- OTPRSweights %>% select(gene, gene_name, effect_size, pvalue)

osobservedexpression_cleaned$gene_name <- row.names(osobservedexpression_cleaned)
osobservedexpression_reference <- osobservedexpression_cleaned 

osobservedexpression_cleaned <- merge(osobservedexpression_cleaned, OTPRSweights, by="gene_name", all.x=TRUE)
osobservedexpression_cleaned <- osobservedexpression_cleaned %>% dplyr::filter(!is.na(effect_size))
osobservedexpression_cleaned <- osobservedexpression_cleaned %>% dplyr::filter(duplicated(gene_name)==FALSE)

osobservedexpressiongenes <- osobservedexpression_cleaned$gene
osobservedexpressionrows <- osobservedexpression_cleaned$gene_name

osobservedexpression_cleaned_t <- osobservedexpression_cleaned %>%
  mutate_at(vars(-gene_name, -gene, -effect_size, -pvalue), function(x) {log(x+1)})

#OTRSscore <- gen.score(observedexpression_cleaned, 647, 2, 645)
OTRSscore <- gen.score(osobservedexpression_cleaned_t, 685, 2, 683)

OTRSscoredev <- OTRSscore
rownames(OTRSscoredev) <- osobservedexpressiongenes

OTRSoptimization <- calculate_trs(OTRSscoredev, OTPRSweights)

pheno <- famOS %>% select(fullID,phenotype)
names(pheno) = c("SUBJID", "phenotype") 

perf_o = list()
otrs <- OTRSoptimization$ptrs
otrs$indiv = stringr::str_replace_all(otrs$indiv, '\\.', '-')
for(c in predixcan_cutoffs) {
  perf_o[[as.character(c)]] = inner_join(pheno, otrs %>% dplyr::filter(cutoff == c), by = c('SUBJID' = 'indiv'))
}

for(i in names(perf_o)) {
    p = perf_o[[i]] %>% ggplot() + geom_density(aes(x = tprs, fill = factor(phenotype)), alpha = .5) + theme(legend.position = 'bottom')
    p = p + ggtitle(paste0('p-value cutoff: ', i))
    ggsave(paste0('~/Desktop/', 'otrs_cutoff_', i, '.pdf'), p, height = 10, width = 6)
  }

OTRSscorefinal <- OTRSoptimization$ptrs %>% filter(cutoff==1e-5)
OTRSscorefinal$indiv = stringr::str_replace_all(OTRSscorefinal$indiv, '\\.', '-')
OTRSscorefinal <- merge(OTRSscorefinal, famOS, by.x="indiv", by.y="fullID")
OTRSscorefinal <- OTRSscorefinal %>% select (indiv, tprs, cutoff, phenotype)
```

```{r Cox regression of Observed TRS and overall survival, cache=TRUE, warning=FALSE, echo=FALSE}
survivalwhiteOTRS <- merge(OTRSscorefinal, TCGAavail, by.x="indiv", by.y="submitter_id")

#Cox regression of overall survival
surv_objectOTRS <- Surv(time=survivalwhiteOTRS$oss.months, event=as.numeric(survivalwhiteOTRS$phenotype))

fit.coxphOTRS <- coxph(surv_objectOTRS ~ survivalwhiteOTRS$tprs,
                   data = survivalwhiteOTRS)

ggforest(fit.coxphOTRS, data = survivalwhiteOTRS, main="Observed transcriptomic score in overall survival\nin European ancestry patients")
```

In a univariate analysis, OTRS is not significantly associated with overall survival in breast cancer. 

# OTRS-Progression-free survival in breast cancer 
```{r Observed-TPRS score generation for progression free survival, cache=TRUE, echo=FALSE, warning=FALSE}
pheno <- famPFS %>% select(fullID,phenotype)
names(pheno) = c("SUBJID", "phenotype") 

perf_o = list()
otrs <- OTRSoptimization$ptrs
otrs$indiv = stringr::str_replace_all(otrs$indiv, '\\.', '-')
for(c in predixcan_cutoffs) {
  perf_o[[as.character(c)]] = inner_join(pheno, otrs %>% dplyr::filter(cutoff == c), by = c('SUBJID' = 'indiv'))
}

for(i in names(perf_o)) {
    p = perf_o[[i]] %>% ggplot() + geom_density(aes(x = tprs, fill = factor(phenotype)), alpha = .5) + theme(legend.position = 'bottom')
    p = p + ggtitle(paste0('p-value cutoff: ', i))
    ggsave(paste0('~/Desktop/', 'otrs_cutoff_', i, '.pdf'), p, height = 10, width = 6)
  }

OTRSscorefinal <- OTRSoptimization$ptrs %>% filter(cutoff==1e-5)
OTRSscorefinal$indiv = stringr::str_replace_all(OTRSscorefinal$indiv, '\\.', '-')
OTRSscorefinal <- merge(OTRSscorefinal, famPFS, by.x="indiv", by.y="fullID")
OTRSscorefinal <- OTRSscorefinal %>% select (indiv, tprs, cutoff, phenotype)
```

```{r Cox regression of Observed TRS and progression-free survival, cache=TRUE, echo=FALSE, warning=FALSE}
recurrencewhiteOTRS <- merge(OTRSscorefinal, TCGAavail, by.x="indiv", by.y="submitter_id")

#Cox regression of recurrence
surv_objectOTRS <- Surv(time=recurrencewhiteOTRS$pfs.months, event=as.numeric(recurrencewhiteOTRS$phenotype))

fit.coxphOTRS <- coxph(surv_objectOTRS ~ recurrencewhiteOTRS$tprs,
                   data = recurrencewhiteOTRS)

ggforest(fit.coxphOTRS, data = recurrencewhiteOTRS, main="Observed transcriptomic score in progression-free survival\nin European ancestry patients")
```

In a univariate analysis, OTRS is not significantly associated with progression-free survival in breast cancer. 