---
title: "Generation of predicted transcriptomic risk score in patients with breast cancer"
author: "Padma Sheila Rajagopal, MD MPH"
date: "1/20/2020"

site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    df_print: paged
    toc: true
    theme: spacelab
    highlight: textmate
    
---
```{r DATA setup, warning=FALSE, include=FALSE, echo=FALSE}
rm(list = ls())
library(tidyverse)
library(dplyr)
library(ggplot2)
library(data.table)
library(pander)
library(ggExtra)
library(survival)
library(survminer)
library(stringr)
library(glmnet)
library(survival)
library(factoextra)
library(flashClust)
knitr::opts_chunk$set(autodep = TRUE)
options(datatable.fread.datatable = F)
options(stringsAsFactors = F)
panderOptions('table.split.table', Inf)
predixcan_cutoffs = c(5e-8, 1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1)

setwd("~/Research-Local/2020-ptrsotrs")
load("Input/PTRSOTRS.RData")

gen.score <- function(df, x, a, b) { 
  ## extract the objective column
  df1 = df[,a:b]
  col = df[, x]
  df1 * col
}

calculate_trs = function(expression_df, effect_size_df, pvalue_cutoffs = c(5e-8, 1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1)) {
  # expression_df should have rownames as gene id and colnames as individual id
  # effect_size_df should have 
  #     1. column named gene with gene id;
  #     2. column named effect_size as effect size;
  #     3. column named pvalue as pvalue
  
  # first of all, we remove genes with crazily large effect size (they will skew the PTRS distribution)
  effect_size_df_cleaned = effect_size_df %>% filter(abs(effect_size) < 5)
  
  # limit analysis to genes occcur in both expression_df and effect_size_df
  pred_expr_gene_pool = rownames(expression_df)
  predixcan_gene_pool = effect_size_df_cleaned$gene
  gene_pool_intersect = intersect(pred_expr_gene_pool, predixcan_gene_pool)
  
  # loop over p-value cutoffs and calculate PTRS at each cutoff
  out = list()
  outgenes = list()
  for(c in pvalue_cutoffs) {
    predixcan_i = effect_size_df_cleaned %>% filter(gene %in% gene_pool_intersect)
    beta_gene = predixcan_i %>% filter(pvalue < c) %>% select(gene, effect_size)
    outgenes[[length(outgenes) + 1]] = beta_gene %>% mutate(cutoff = c)
    indiv = colnames(expression_df) 
    mat = t(expression_df)
    mat = mat[, colnames(mat) %in% beta_gene$gene, drop = F]
    mat = mat[, match(beta_gene$gene, colnames(mat)), drop = F]
    tprs_i = as.matrix(mat) %*% beta_gene$effect_size
    df_tprs = data.frame(indiv = indiv, tprs = tprs_i, cutoff = c)
    rownames(df_tprs) = NULL
    out[[length(out) + 1]] = df_tprs
  }
  out = do.call(rbind, out)
  outgenes = do.call(rbind, outgenes)
  list(ptrs = out, selected_genes = outgenes)
}
```

# Predicted transcriptomic score (PTRS)
This is the code to generate the predicted transcriptomic risk score and assess its ability to prognosticate overall survival and progression free survival among breast cancer patients. 

The predicted transcriptomic score is derived from using the weight of a gene's association to germline risk of breast cancer * predicted expression of that gene, summed across different thresholds of signifcance. 

```{r Predicted-TPRS score generation for overall survival, cache=TRUE, warning=FALSE, echo=FALSE}
setwd("~/Research-Local/2020-ptrsotrs")
ospredictedexpression <- read.table("Input/predicted_expression/BRCA/os-results_predicted_expression.txt", header=TRUE, as.is = TRUE)
ospredictedexpressionreference <- ospredictedexpression
PTRSOSgroup <- ospredictedexpressionreference$IID

PTPRSweights <- read.table("Input/risk_weights/2020_BRCA_risk_weights/BCAC_Overall_2020_spredixcan_GTExV8_mashr.csv", header=TRUE, sep=",")
PTPRSweights <- PTPRSweights %>% dplyr::select(gene, gene_name, effect_size, pvalue)

ospredictedexpression <- as.data.frame(t(ospredictedexpression))
names(ospredictedexpression) <- ospredictedexpression["IID",]

ospredictedexpression <- ospredictedexpression[-1,] #Removing the FID row
ospredictedexpression <- ospredictedexpression[-1,] #Removing the IID row 
ospredictedexpressiongene <- rownames(ospredictedexpression)
ospredictedexpression <- as.data.frame(lapply(ospredictedexpression, as.numeric))
rownames(ospredictedexpression) <- ospredictedexpressiongene
ospredictedexpression$gene <- row.names(ospredictedexpression)

ospredictedexpression <- merge(ospredictedexpression, PTPRSweights, by="gene", all.x=TRUE)

PTRSscoreraw <- gen.score(ospredictedexpression, 697, 2, 695)
PTRSscoredev <- PTRSscoreraw
rownames(PTRSscoredev) <- ospredictedexpressiongene

PTRSoptimization <- calculate_trs(PTRSscoredev, PTPRSweights)
pheno <- famOS %>% select(fullID,phenotype)
names(pheno) = c("SUBJID", "phenotype") 

perf_t = list()
ptrs <- PTRSoptimization$ptrs
ptrs$indiv = stringr::str_replace_all(ptrs$indiv, '\\.', '-')
for(c in predixcan_cutoffs) {
  perf_t[[as.character(c)]] = inner_join(pheno, ptrs %>% dplyr::filter(cutoff == c), by = c('SUBJID' = 'indiv'))
}

for(i in names(perf_t)) {
    p = perf_t[[i]] %>% ggplot() + geom_density(aes(x = tprs, fill = factor(phenotype)), alpha = .5) + theme(legend.position = 'bottom')
    p = p + ggtitle(paste0('p-value cutoff: ', i))
    ggsave(paste0('~/Desktop/', 'ptrs_cutoff_', i, '.pdf'), p, height = 10, width = 6)
  }

PTRSscorefinal <- PTRSoptimization$ptrs %>% dplyr::filter(cutoff==1e-5)
PTRSscorefinal$indiv = stringr::str_replace_all(PTRSscorefinal$indiv, '\\.', '-')
PTRSscorefinal <- merge(PTRSscorefinal, famOS, by.x="indiv", by.y="fullID")
PTRSscorefinal <- PTRSscorefinal %>% select (indiv, tprs, cutoff, phenotype)
```

# PTRS-Overall survival in breast cancer 
```{r Cox regression of Predicted TRS and overall survival, cache=TRUE, warning=FALSE, echo=FALSE}
survivalwhitePTRS <- merge(PTRSscorefinal, TCGAavail, by.x="indiv", by.y="submitter_id")
survivalwhitePTRS$phenotype <- ifelse(survivalwhitePTRS$phenotype=="2", "case", survivalwhitePTRS$phenotype)
survivalwhitePTRS$phenotype <- ifelse(survivalwhitePTRS$phenotype=="1", 0, survivalwhitePTRS$phenotype)
survivalwhitePTRS$phenotype <- ifelse(survivalwhitePTRS$phenotype=="case", 1, survivalwhitePTRS$phenotype)

#Cox regression of overall survival
surv_objectPTRS <- Surv(time=survivalwhitePTRS$oss.months, event=as.numeric(survivalwhitePTRS$phenotype))

fit.coxphPTRS <- coxph(surv_objectPTRS ~ survivalwhitePTRS$tprs,
                   data = survivalwhitePTRS)

ggforest(fit.coxphPTRS, data = survivalwhitePTRS, main="Predicted transcriptomic score in overall survival in European ancestry patients")
```

In a univariate analysis, PTRS does demonstrate a significant assoication with overall survival (OS). Of note, the direction of effect is not as expected, and the threshold of significance is borderline. This score incorporates the germline in both genes associated with risk (by weights) and in terms of the predicted transcriptome. 

# Control for Predicted-TPRS score - overall transcriptome and overall survival
```{r Control.1 for Predicted-TPRS score - overall transcriptome without risk-based effect size and overall survival, cache=TRUE, warning=FALSE, echo=FALSE}
ospredictedexpressioncontrol1 <- ospredictedexpressionreference
ospredictedexpressioncontrol1 <- as.data.frame(t(ospredictedexpressioncontrol1))
names(ospredictedexpressioncontrol1) <- ospredictedexpressioncontrol1["IID",]
ospredictedexpressioncontrol1 <- ospredictedexpressioncontrol1[-1,] #Removing the FID row
ospredictedexpressioncontrol1 <- ospredictedexpressioncontrol1[-1,] #Removing the IID row 
ospredictedexpressioncontrol1gene <- rownames(ospredictedexpressioncontrol1)
ospredictedexpressioncontrol1 <- as.data.frame(lapply(ospredictedexpressioncontrol1, as.numeric))
rownames(ospredictedexpressioncontrol1) <- ospredictedexpressioncontrol1gene
ospredictedexpressioncontrol1$gene <- row.names(ospredictedexpressioncontrol1)

ospredictedexpressioncontrol1$weight <- 1
PTRScontrolraw <- gen.score(ospredictedexpressioncontrol1, 696, 1,694)

PTRScontroldev <- PTRScontrolraw
rownames(PTRScontroldev) <- ospredictedexpressioncontrol1gene

PTRScontrol1 <- as.data.frame(colSums(PTRScontroldev, na.rm = FALSE))
colnames(PTRScontrol1) <- c("PTRScontrol1")
PTRScontrol1$SUBJID <- rownames(PTRScontrol1)
PTRScontrol1$SUBJID = stringr::str_replace_all(PTRScontrol1$SUBJID, '\\.', '-')

pheno <- famOS %>% select(fullID,phenotype)
names(pheno) = c("SUBJID", "phenotype") 

PTRScontrol1 <- merge(PTRScontrol1, pheno, by="SUBJID")
colnames(PTRScontrol1) <- c("indiv", "ptrscontrol", "phenotype")
```

```{r Cox regression of Control.1 for predicted TPRS score (overall transcriptome without risk-based effect size) and overall survival, cache=TRUE, warning=FALSE, echo=FALSE}

survivalwhitePTRScontrol <- merge(PTRScontrol1, TCGAavail, by.x="indiv", by.y="submitter_id")

survivalwhitePTRScontrol$phenotype <- ifelse(survivalwhitePTRScontrol$phenotype=="2", "case", survivalwhitePTRScontrol$phenotype)
survivalwhitePTRScontrol$phenotype <- ifelse(survivalwhitePTRScontrol$phenotype=="1", 0, survivalwhitePTRScontrol$phenotype)
survivalwhitePTRScontrol$phenotype <- ifelse(survivalwhitePTRScontrol$phenotype=="case", 1, survivalwhitePTRScontrol$phenotype)

#Cox regression of overall survival
surv_objectPTRScontrol <- Surv(time=survivalwhitePTRScontrol$oss.months, event=as.numeric(survivalwhitePTRScontrol$phenotype))

fit.coxphPTRScontrol <- coxph(surv_objectPTRScontrol ~ survivalwhitePTRScontrol$ptrscontrol,
                   data = survivalwhitePTRScontrol)

ggforest(fit.coxphPTRScontrol, data = survivalwhitePTRScontrol, main="Predicted transcriptomic score in overall survival in European ancestry patients (all genes)\n - no risk weights")
```

In a univariate analysis, just using the germline predicted expression values without the weighting based on risk is not significantly associated with OS. 

# PTRS-Progression-free survival in breast cancer 
```{r Predicted-TPRS score generation for progression-free survival, cache=TRUE, echo=FALSE, warning=FALSE}
setwd("~/Research-Local/2020-ptrsotrs")
pfspredictedexpression <- read.table("Input/predicted_expression/BRCA/pfs-results_predicted_expression.txt", header=TRUE, as.is = TRUE)
pfspredictedexpressionreference <- pfspredictedexpression

pfspredictedexpression <- as.data.frame(t(pfspredictedexpression))
names(pfspredictedexpression) <- pfspredictedexpression["IID",]

pfspredictedexpression <- pfspredictedexpression[-1,] #Removing the FID row
pfspredictedexpression <- pfspredictedexpression[-1,] #Removing the IID row 
pfspredictedexpressiongene <- rownames(pfspredictedexpression)
pfspredictedexpression <- as.data.frame(lapply(pfspredictedexpression, as.numeric))
rownames(pfspredictedexpression) <- pfspredictedexpressiongene
pfspredictedexpression$gene <- row.names(pfspredictedexpression)

pfspredictedexpression <- merge(pfspredictedexpression, PTPRSweights, by="gene", all.x=TRUE)

PTRSscoreraw <- gen.score(pfspredictedexpression, 697, 2, 695)
PTRSscoredev <- PTRSscoreraw
rownames(PTRSscoredev) <- pfspredictedexpressiongene

PTRSoptimization <- calculate_trs(PTRSscoredev, PTPRSweights)

pheno <- famOS %>% select(fullID,phenotype)
names(pheno) = c("SUBJID", "phenotype") 

perf_t = list()
ptrs <- PTRSoptimization$ptrs
ptrs$indiv = stringr::str_replace_all(ptrs$indiv, '\\.', '-')
for(c in predixcan_cutoffs) {
  perf_t[[as.character(c)]] = inner_join(pheno, ptrs %>% dplyr::filter(cutoff == c), by = c('SUBJID' = 'indiv'))
}

for(i in names(perf_t)) {
    p = perf_t[[i]] %>% ggplot() + geom_density(aes(x = tprs, fill = factor(phenotype)), alpha = .5) + theme(legend.position = 'bottom')
    p = p + ggtitle(paste0('p-value cutoff: ', i))
    ggsave(paste0('~/Desktop/', 'ptrs_cutoff_', i, '.pdf'), p, height = 10, width = 6)
  }

PTRSscorefinal <- PTRSoptimization$ptrs %>% dplyr::filter(cutoff==1e-5)
PTRSscorefinal$indiv = stringr::str_replace_all(PTRSscorefinal$indiv, '\\.', '-')
PTRSscorefinal <- merge(PTRSscorefinal, famPFS, by.x="indiv", by.y="fullID")
PTRSscorefinal <- PTRSscorefinal %>% select (indiv, tprs, cutoff, phenotype)
```

```{r Cox regression of Predicted TRS and recurrence, cache=TRUE, warning=FALSE, echo=FALSE}
recurrencewhitePTRS <- merge(PTRSscorefinal, TCGAavail, by.x="indiv", by.y="submitter_id")
recurrencewhitePTRS$phenotype <- ifelse(recurrencewhitePTRS$phenotype=="2", "case", recurrencewhitePTRS$phenotype)
recurrencewhitePTRS$phenotype <- ifelse(recurrencewhitePTRS$phenotype=="1", 0, recurrencewhitePTRS$phenotype)
recurrencewhitePTRS$phenotype <- ifelse(recurrencewhitePTRS$phenotype=="case", 1, recurrencewhitePTRS$phenotype)

#Cox regression of recurrence
surv_objectPTRS <- Surv(time=recurrencewhitePTRS$pfs.months, event=as.numeric(recurrencewhitePTRS$phenotype))

fit.coxphPTRS <- coxph(surv_objectPTRS ~ recurrencewhitePTRS$tprs,
                   data = recurrencewhitePTRS)

ggforest(fit.coxphPTRS, data = recurrencewhitePTRS, main="Predicted transcriptomic score in progression-free survival in European ancestry patients")
```

In a univariate analysis, PTRS is not significantly associated with progression-free survival in breast cancer. 